```{r setup, cache = F, echo = F}
library(knitr)
opts_chunk$set(tidy = F,  fig.align = "center", fig.height = 5, fig.width = 10)
```

# Reproducible Research

We saw `R` is a powerful software for data analysis. Contrary to other softwares
which provide you with ready, fixed, *canned* procedures to perform statistical
analyses, R always let you have access to all computations' results. R can also be
extended with *packages*. About 6,000 of them are available via
[CRAN](http://cran.r-project.org/web/packages/). Furthermore, you can easily
create your own packages, or functions, to tackle your specific needs and/or
problems.

Specifically for epidemiologist, you could use the followings packages:

- [Epi](http://cran.r-project.org/web/packages/Epi/index.html) (DAN)
- [epibasix](http://cran.r-project.org/web/packages/epibasix/index.html) (CAN)
- [epicalc](http://cran.r-project.org/web/packages/epicalc/index.html) (THA)
- [epiR](http://cran.r-project.org/web/packages/epiR/index.html) (NZE, vet)
- [epitools](http://cran.r-project.org/web/packages/epitools/index.html) (USA)
-
  [EpiContactTrace](http://cran.r-project.org/web/packages/EpiContactTrace/index.html)
-(SWE)
- [EpiEstim](http://cran.r-project.org/web/packages/EpiEstim/index.html) (UK)
- [EpiModel](http://cran.r-project.org/web/packages/EpiModel/index.html) (USA)

All good but you have to be able to reproduce your analysis, redo your figures
and tables i.e. have availabe te computer code which produce your research
results (more on that for example in
[Peng et al.](http://aje.oxfordjournals.org/content/163/9/783)). 

We will use the [pig_adg](http://www.upei.ca/ver/datasets-programs) dataset from
the [VER](http://www.upei.ca/ver/welcome) book by Ian Dohoo et al. These are
data on growth performance and abattoir findings of pigs, studying the
inter-relationships among respiratory diseases (atrophic rhinitis and enzootic
pneumonia), ascarid levels and daily weight gain
([Bernardo, Dohoo, and Donald](http://www.ncbi.nlm.nih.gov/pmc/articles/PMC1255650/)). The
book website offers dataset in R format.

```{r read-data}
temp <- tempfile()
download.file(
    "http://ic.upei.ca/ver/sites/ic.upei.ca.ver/files/ver2_data_R.zip", temp)
load(unz(temp, "ver2_data_R/pig_adg.rdata"))
unlink(temp)

pig_adg$ar_g1 <- with(pig_adg, ifelse(ar > 1, 1, 0))
```

## Saving Figures to File

`R` can generate graphics in any type of display (pdf, postscript, bitmap, svg,
png, jpeg, tiff). Just tell `R` to save the current image displayed:

```{r, eval = FALSE}
plot(1:10)
dev.copy(png, file = "myPlot.png")
dev.off()
```

If you want to save the plot without seeing it *on screen*:

```{r, eval = FALSE}
pdf(file = "myPlot.pdf")
plot(1:10)
dev.off()
```

If your plot was made with `ggplot2`, you use:

```{r, eval = FALSE}
ggsave("myPlot.pdf")
```

## Saving Commands in R Scripts

You save your commands in an `R` script (e.g. `myScript.R`). You can then go
through it line by line.

## Sourcing Code

`source()` runs `R` commands held in an external source file, or `R` script. If
you set the option to `TRUE`, it will print the commands and results to the
console.

```{r, eval = FALSE}
source('myScript.R')
```

You can also `sink()` or `capture.output` selected results from your source file
to an output file. For example if our source file produce summary statistcs from
`pig_adg`:

```{r, eval = FALSE}
pig.sum <- summary(pig_adg)
sink("summary.log")
cat("My summary stats", fill = TRUE)
show(pig.sum)
sink()
```

```{r, eval = FALSE}
pig.sum <- summary(pig_adg)
capture.output({pig.sum}, file = "summary.txt")
```

You can even source several `R` scripts into your main source file, like functions
you wrote.

## Source File from Shell

Next, you can also run your script from the command line. For example save the
following as an `R` script:

```{r, eval = FALSE}
temp <- tempfile()
download.file(
    "http://ic.upei.ca/ver/sites/ic.upei.ca.ver/files/ver2_data_R.zip", temp)
load(unz(temp, "ver2_data_R/pig_adg.rdata"))
unlink(temp)

pig_adg$ar_g1 <- with(pig_adg, ifelse(ar > 1, 1, 0))

library(epicalc)
with(pig_adg, cc(pn, ar_g1,
   decimal = 2,
   graph = FALSE, 
   original = TRUE,
   design = "cohort",
   alpha = .05,
   fisher.or = FALSE, 
   exact.ci.or = TRUE))

with(pig_adg, mhor(pn, ar_g1, sex,
               design = "cohort",
               decimal = 2,
               graph = FALSE))
```

Now you can call it from the command line and save its output with `Rscript
yourScript.R > yourOutput.Rout`. You can also use the batch command `R CMD BATCH
yourScript.R yourOutput.txt`. Contrary to the file redirection, `R CMD BATCH`
captures warnings and errors. `R CMD BATCH` also adds a call to `proc.time()` to
give the time the script took to run.

## Stitch and Spin

Better yet is to embed the results and the code that generated it in the same
document. For this we use [`knitr`](http://yihui.name/knitr/). In your script,
everything written after a comment `#` will be converted to text and everything
else will be assumed to be `R` code and executed with the `spin()` function. A
[markdown](https://en.wikipedia.org/wiki/Markdown) file is generated. It then
can be converted into a MS Office document with
[pandoc](http://johnmacfarlane.net/pandoc/). Try with the following and call it
with `Rscript -e "knitr::spin('yourScript.R')"`. Then to convert it to
docx: `pandoc -s -S yourScript.md -o results.docx` (note: you might not
have `pandoc` installed on your computer).

```{r, eval = FALSE}
#' Pig Pneumonia

#+ echo=FALSE
temp <- tempfile()
download.file(
    "http://ic.upei.ca/ver/sites/ic.upei.ca.ver/files/ver2_data_R.zip", temp)
load(unz(temp, "ver2_data_R/pig_adg.rdata"))
unlink(temp)
pig_adg$ar_g1 <- with(pig_adg, ifelse(ar > 1, 1, 0))

#' The results of the logistic regression are the following:

#+ log, echo=FALSE
mod1 <- glm(pn ~ ar_g1,
             family = binomial("logit"),
             data = pig_adg)  # "logit" can be omitted as it is the default

#+ log-sum
summary(mod1)
```

You can also use `stitch()` which insert the `R` code into a predefined template
and create a simple report (`Rscript -e "knitr::stitch_rhtml(yourScript.R')`):

```{r, eval = FALSE}
temp <- tempfile()
download.file(
    "http://ic.upei.ca/ver/sites/ic.upei.ca.ver/files/ver2_data_R.zip", temp)
load(unz(temp, "ver2_data_R/pig_adg.rdata"))
unlink(temp)
pig_adg$ar_g1 <- with(pig_adg, ifelse(ar > 1, 1, 0))

library(lme4)
mod2 <- glmer(pn ~ ar_g1 + (1|farm),
              family = binomial,
              data = pig_adg)
summary(mod2)

se <- sqrt(diag(vcov(mod2)))
# table of estimates with 95% CI
(tab <- cbind(Est = fixef(mod2),
              LL = fixef(mod2) - 1.96 * se,
              UL = fixef(mod2) + 1.96 * se))
exp(tab)
```

## Literate programming with `knitr`

**Basic idea:** Write **data** + **software** + **documentation** (or in this
  case manuscripts, reports) together.

Analysis code can be divided into text and code "chunks". Doing so allows us to
extract the code for machine readable documents (technically referred to as a
`tangle`) or produce a human-readable document (also called `weave`).

Literate programming involves three main steps:  

1. Separate the narrative from the code
2. Execute source code and return the results.
3. Combine the results from the source code with the original narratives to
   produce a final document.

### Why this is important?
Results from scientific research have to be easy to reproduce so others can
verify results making them trustworthy. Otherwise we risk producing one off
results that no one outside the original research group can reproduce. In this
lesson we will learn reproducible research, which is one of the by products of
dynamics report generation. However, this process alone will not always
guarantee reproducibility.

### Installing `knitr`

```coffee
# Installing knitr is quite easy. 
install.packages("knitr", dependencies = TRUE)
install.packages("pander") # Pander is a useful package for formatting tables in markdown.
```

Knitr supports a variety of documentation formats including `markdown`, `html`
and `LaTeX`. It also allows for easy export to `PDF` and `HTML`.

### What is markdown?

Markdown is an incredibly simple semantic file format, not too dissimilar from
.doc, .rtf, or .txt. Markdown makes it easy for even those without significant
knowledge of markup languages like html or LaTex to write any sort of text
(including with links, lists, bullets, etc.) and have it parsed into a variety
of formats.

* [Markdown cheatsheet](https://github.com/adam-p/markdown-here/wiki/Markdown-Cheatsheet)
* [Original markdown reference](http://daringfireball.net/projects/markdown/basics)

### When to implement reproducibility via literate programming?

* You can do this anytime but it is always best to do this at the beginning of a
  project. Much like writing unit tests, waiting till after a project is
  completed is a bad idea. 
* Best used alongside version control like Git. 
* Use software like R where instructions are scripted.
* Never save the output (only raw dataset with pre-processing code. Even cleaned
  datasets can be discarded but it might help to save them temporarily during
  intermediate  steps).
* Finally, store the data in a non-proporietary format (e.g. `csv` over
  `xls`). This will ensure that your data will be readable long into the future. 

---

**Good and Bad Practices**

Related: See [best practices](../R-basics/best-practices.Rmd) in the
[R-basics folder](../R-basics/).

### Creating a basic knitr document

In RStudio, choose new R Markdown file (easiest way) or you can create a new
text file and save it with extension `.Rmd`.

A basic code chunk looks like this:

```{r, eval = FALSE}
# some R code
```

---

You can knit this document using the knit button or do it programmatically using
the `knit()` function.

```{r, eval = FALSE}
library(knitr)
knit("file.Rmd")
```

**What just happened?**

knitr read the Rmd file, then located and ran all the code chunks identified by
the backticks, and replaced it with the output of R function calls. If figures
are generated from any such calls, they will be included in markdown syntax.  

### Chunk labels

You can also name your code chunks. This allows you to keep all the code in a
separate script and just refer to code chunks using meaningful names
(e.g. data-processing, analysis, model-fitting, visualization, figures, tables)

```{r, eval = FALSE}
{r, chunk_name}
```


__Some rules on naming chunks__
* Chunk labels are supposed to be unique id’s in a document.  
* knitr will throw an error if two chunks have the same name.  
* If no chunk names are given, knitr will simply increment from chunk 1, 2,3 etc.

In addition to naming chunks within the curly braces, you can also add a bunch
of other options on how that particular code chunk should behave. 

**Other options you can add to the tag**

| Option | Description |
| ------ | ------------ |
| **echo** =   TRUE or FALSE |  to show or hide code.  |
| **eval** =   TRUE or FALSE | to run or skip the code.  |
| **warning** =   TRUE or FALSE | to show or hide function warnings.  |
| **message** =  TRUE or FALSE | to show or hide function R messages.  |
| **results** = "hide"  | will hide results. They will still be executed |
| **fig.height** = | Height of figure  |
| **fig.width** =  | width of figure  |

Once your output markdown files (`.md`) files are generated, you should never
edit them because they are automatically generated. Next time you knit the
original `.Rmd` files, all the changes in the `.md` file will get wiped out. 

**Write sentences in text with inline output**

```{r, eval = FALSE}
Include some text `r mean(1:5)`. 
```

### Including formatted tables in markdown

```{r showtable, results="asis", echo = FALSE, message = FALSE, eval = FALSE, warning = FALSE}
library(pandoc)
pander(summary(mod2))
```

### Global options

Global options are shared across all the following chunks after the location in
which the options are set, and local options in the chunk header can override
global options.

```{r setoptions, eval = FALSE, echo = FALSE}
options(width = 60, show.signif.stars = FALSE)
opts_chunk$set(echo = FALSE, 
            results = "asis", 
            warning = FALSE, 
            message = FALSE, 
            fig.width = 5, 
            fig.height = 4, 
            tidy = TRUE, 
            fig.align = 'center')
```

### Other Options
**Dealing with long running process**

By adding `cache = TRUE` to a code block definition. After the first run,
results will be cached. We'll discuss better ways to acheive the same thing
using `Make` in the next section.

### Working with graphics in `knitr`

If you use `ggplot2` from the data visualization section, you can have that
easily parsed into your document.

```{r, eval = FALSE}
library(ggplot2)
p <- qplot(carat, price, data = diamonds) + geom_hex()
p 
# no need to explicitly print(p)
```

---

### Pandoc
Pandoc (http://johnmacfarlane.net/pandoc) is a universal document converter. In
particular, Pandoc can convert Markdown to many other document formats,
including LATEX, HTML, Rich Text Format (*.rtf), E- Book (*.epub), Microsoft
Word (*.docx) and OpenDocument Text (*.odt), etc.
Pandoc is a command line tool. Linux users should be fine with it; for Windows
users, the command window can be accessed via the Start menu, then Run cmd. Once
we have opened a command window (or terminal), we can type commands like this to
convert a Markdown file, say, test.md, to other formats:

```
pandoc test.md -o test.html
pandoc test.md -o test.odt
pandoc test.md -o test.rtf
pandoc test.md -o test.docx
pandoc test.md -o test.pdf
```

You can add more options to the basic Pandoc call. To see a full list of options

```
pandoc --help
```

A commonly used option is to add margins using the `-V` argument (in this case 1
inch):

```
pandoc -V geometry:margin=1in test.md -o test.pdf
```
